---
title: "Campbellgram - joint visualisation of copy number and structural variations"
author: "Yilong Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Campbellgram - joint visualisation of copy number and structural variations using campbellgrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width=7,
    fig.height=4,
    dpi=300,
    out.width="700px",
    out.height="400px"
)
```

Campbellgram is an R package for visualising *absolute* copy number data and structural variation data jointly in the same plot. The basic premise of the package is its simplicity of use: all required inputs of the main function, `campbellgram()`, are standard R data frames.

In addition, `campbellgram()` allows a mutation rainfall plot to be overlaid on top of a campbellgram.

## Plotting campbellgrams

The simplest usage of `campbellgram()` involves plotting nothing but structural variations. In this case, the only required arguments are

* `bedpe` - a data frame of structural variation junctions in [the BEDPE format](https://bedtools.readthedocs.io/en/latest/content/general-usage.html#bedpe-format). Row and column names are ignored. The ID and score columns (columns 7 and 8) are also ignored, although they need to be present in the data frame for it to conform to the BEDPE format.
* `cn_bedgraph` - a data frame of ***absolute copy number data** in [the bedGraph format](https://genome.ucsc.edu/goldenPath/help/bedgraph.html). Row and column names are ignored.
* `chrs_used` - a vector of chromosomes to be visualised, and
* `ref_genome` - a data frame with chromosomes name and chromosome length on its first and second column, respectively.

Data for `ref_genome` can be generated by using [`samtools faidx`](http://www.htslib.org/doc/samtools.html). Alternatively, function `fetchExtendedChromInfoFromUCSC` from package [`GenomeInfoDb`](https://bioconductor.org/packages/release/bioc/html/GenomeInfoDb.html) can be used to fetch the chromosome length data online.

```{r}
library(campbellgram)
data(example_bedpe)
knitr::kable(head(example_bedpe))
data(example_cn)
knitr::kable(head(example_cn))
data(hs37d5_chrom_sizes)
knitr::kable(head(hs37d5_chrom_sizes))
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = c("7", "12"),
             cn_bedgraph = example_cn)
```

Intra-chromosomal structural variations are shown as one of four different types of arcs. Arcs emanating from the first dashed line are deletion ("+-")-type and tandem duplication ("-+")-type SVs. Arcs emanating from the second dashed line are head-to-head ("--") and tail-to-tail ("++") type SVs. The default colours of the arcs are taken from `ARC_COLS`.

```{r}
campbellgram::ARC_COLS
```

## Inter vs intra-chromosomal structural variations

In the terminology of `campbellgram()`, structural variation junctions within any chromosome listed in `chrs_used` are considered intra-chromosomal and displayed using arcs. In contrast, inter-chromosomal structural variation breakpoints are indicated using black vertical lines that end in a left or right-leaning hinge for "+" or "-" orientation breakpoints.

For instance, if the chromosomes 7 and 12 shown above are plotted separately, all the structural variation junctions between chromosomes 7 and 12 will now be displayed as inter-chromosomal translocations.

```{r}
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = "7",
             cn_bedgraph = example_cn)
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = "12",
             cn_bedgraph = example_cn)
```

## Zooming

To zoom to a specific region in a chromosome, use parameter `xlim`.

```{r}
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = "7",
             cn_bedgraph = example_cn, xlim = c(70e6, 100e6))
```

It is also possible to zoom into a chromosome and a chromosomal position while multiple chromosomes are plotted, i.e. when `chrs_used` contains multiple chromosomes. To achieve this, use `chrs_shown` to choose the chromosome to be shown and `xlim` to limit the base pairs further.

```{r}
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = c("7", "12"),
             cn_bedgraph = example_cn, chrs_shown = "7")
campbellgram(hs37d5_chrom_sizes, example_bedpe, chrs_used = c("7", "12"),
            cn_bedgraph = example_cn, chrs_shown = "7", xlim = c(70e6, 100e6))
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output:
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side.

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
